\section{Euclidean algorithm -- GCD}

Now let me use the Euclidean property to compute the gcd of two integers.

Let's use the division algorithm on $20$ and $6$.
\[
20 = 6 \cdot 3 + 2, \,\,\,\,\, 0 \leq 2 < 6
\]
Suppose I want to compute $\gcd(20, 6)$.
Of course the example is small enough that we know that it is $2$.
But notice something about this:
\[
20 = 6 \cdot 3 + 2, \,\,\,\,\, 0 \leq 2 < 6
\]
If $d$ is a divisor of $20$ and $6$, then it must also divide
$2$.
Therefore $\gcd(20, 6)$ must divide $2$.
The converse might not be true.
In general, we have this crucial
bridge between Euclidean property and 
common divisors:

\begin{lem}
  \textnormal{(\defone{GCD Lemma})}
  If $a,b,q,r \in \Z$ such that  
  \[
  a = bq + r
  \]
  then
  \[
  \{ d \mid \text{d is a common divisor of $a, b$} \}
  =
  \{ d \mid \text{d is a common divisor of $b, r$} \}
  \]
  Hence
  \[
  \gcd(a,b) = \gcd(b, r)
  \]
\end{lem}
\proof
TODO


In particular, 
given $a, b \in \Z$ where $a > b > 0$.
By the Euclidean property of $\Z$, there exist $q,r \in \Z$ such that
\[
a = bq + r, \,\,\, 0 \leq r < b
\]
Hence
\[
\gcd(a,b) = \gcd(b, r)
\]

Note that in the above, I only require $a = bq + r$.
For instance for to $\gcd(120, 15)$, I can use
$120 = 1\cdot 15 + (120 - 15)$, i.e., $a = 120, b = 15, q = 1, r = 120-15$.
Then $\gcd(120, 15) = \gcd(15, 120-15) = \gcd(15, 105)$.

However if I use the division algorithm, then 
$r$ is \lq\lq small'':
\[
0 \leq r < b
\]
So if you want to compute $\gcd(a,b)$, make sure $a \geq b$ (otherwise
swap them).
Then $\gcd(a, b) = \gcd(b, r)$ and you would have $a \geq b > r$.
So instead of computing $\gcd(a, b)$, you are better off
computing $\gcd(b, r)$.

But like I said, we do not need the $q$ and $r$ to be the quotient
and remainder.
For instance suppose I want to compute the GCD of 514 and 24.
\[
514 = 24 \cdot 1 + (514 - 24)
\]
Then 
\[
\gcd(514, 24) = \gcd(24, 514 - 24)
\]
which gives us
\[
\gcd(514, 24) = \gcd(24, 490)
\]

Note that $\gcd(0, n) = n$ for any positive integer $n$.
I'll let you think about that one.
(Remember what I said before: 0 is in some sense a big number,
like a black hole. Because every positive number divides $0$.)

Of course this gives rise to the following algorithm
\begin{Verbatim}[frame=single,fontsize=\footnotesize]
ALGORITHM: GCD
Inputs: a, b
Output: gcd(a, b)

if b > a:
    swap a, b

if b == 0:
    return a
else:
    return GCD(a - b, b) 
\end{Verbatim}

This only subtracts one copy of $b$ from $a$.
Suppose we can compute
\[
a = bq + r, \,\,\,\,\, 0 \leq r < b
\]
Then
\[
\gcd(a, b) = \gcd(b, r) 
\]
Of course $r$ is the remainder when $a$ is divided by $b$.
Using this we rewrite the above code to get
the
\defterm{Euclidean Algorithm}\index{Euclidean Algorithm}\tinysidebar{Euclidean Algorithm}:
\begin{Verbatim}[frame=single,fontsize=\footnotesize]
ALGORITHM: GCD (Euclidean algorithm)
Inputs: a, b
Output: gcd(a, b)

if b > a:
    # To make sure that for gcd(a,b), a >= b
    swap a, b

if b == 0: 
    return a
else:
    return GCD(b, a % b)
\end{Verbatim}
Note that if \verb!a! $<$ \verb!b!, then
\[
\texttt{GCD(a, b) = GCD(b, a \% b) = GCD(b, a)}
\]
Therefore the swap is not necessary:
\begin{Verbatim}[frame=single,fontsize=\footnotesize]
ALGORITHM: GCD (Euclidean algorithm)
Inputs: a, b
Output: gcd(a, b)

if b == 0: 
    return a
else:
    return GCD(b, a % b)
\end{Verbatim}
In this case, I'm assuming that \verb!a % b! is available.
As an example:
\begin{align*}
  \gcd(514, 24)
  &= \gcd(24, 514 \% 24) = \gcd(24, 10) \\
  &= \gcd(10, 24 \% 10) = \gcd(10, 4) \\
  &= \gcd(10, 10 \% 4) = \gcd(10, 2) \\
  &= \gcd(2, 10 \% 2) = \gcd(2, 0) \\
  &= 2
\end{align*}

The above can also be done in a loop:
\begin{Verbatim}[frame=single,fontsize=\footnotesize]
ALGORITHM: GCD (Euclidean algorithm)
Inputs: a, b
Output: gcd(a, b)

while 1:
    if b == 0: 
        return a
    else:
        a, b = b, a % b
\end{Verbatim}

\begin{ex}
  Compute the following using the Euclidean Algorithm explicitly.
  \begin{enumerate}[nosep,label=\textnormal{(\alph*)}]
  \item $\gcd(10, 1)$
  \item $\gcd(10, 10)$
  \item $\gcd(107, 5)$
  \item $\gcd(107, 26)$
  \item $\gcd(84, 333)$
  \end{enumerate}
\end{ex}

\begin{ex}
  Compute the following.
  You should go as far as you can.
  In other words, either you can a fixed integer (such as $1$) or
  derive the $\gcd(\alpha, \beta)$ where $\alpha, \beta$ are
  as simple as possible.
  For instance, to simplify $\gcd(3 + 2a, a)$, since $3 + 2a = 2 \cdot a + 3$,
  we have
  \[
  \gcd(3 + 2a, a) = \gcd(a, 3)
  \]
  In the following $a, b, x, n \in \Z$ are positive integers.
  \begin{enumerate}[nosep,label=\textnormal{(\alph*)}]
  \item $\gcd(ab, b)$
  \item $\gcd(a, a + 1)$
  \item $\gcd(ab + a, b)$ where $0 < a < b$
  \item $\gcd(a(a+1) + a, (a+1))$ where $0 < a < b$
  \item $\gcd(1 + x + \cdots + x^n, x)$
  \item $\gcd(F_{10}, F_{11})$ where $F_n$ is the $n$--th Fibonacci number.
    (Recall: $F_0 = 1, F_1 = 1, F_{n + 2} = F_{n + 1} + F_n$ for $n \geq 0$.)
  \end{enumerate}
\end{ex}

Despite the fact that the Euclidean algorithm is one of the fastest
algorithm to compute the GCD of two numbers and has been discovered
by
\href{https://en.wikipedia.org/wiki/Euclid}{Euclid} a long time ago (BC 300),
the actual runtime was not known until
\href{https://en.wikipedia.org/wiki/Gabriel_Lam%C3%A9}{Lam\'e}
proved in 1844 that the number of steps to compute
$\gcd(a, b)$ using the Euclidean algorithm
is $\leq 5$ times the number of digits (in base 10 notation)
of $\min(a, b)$.
For instance for the example above
of $\gcd(514, 24)$, the number of digits of $\min(514,24)$ is $2$.
Lam\'e theorem says that the number of steps made by the
Euclidean algorithm in the computation of $\gcd(514, 24)$ is
at most $5 \times 2 = 10$.
The actual number of steps in the earlier computation
\begin{align*}
  \gcd(514, 24)
  &= \gcd(24, 514 \% 24) = \gcd(24, 10) \\
  &= \gcd(10, 24 \% 10) = \gcd(10, 4) \\
  &= \gcd(10, 10 \% 4) = \gcd(10, 2) \\
  &= \gcd(2, 10 \% 2) = \gcd(2, 0) \\
  &= 2
\end{align*}
is 4 (not counting the base case step), i.e.,
\[
  \gcd(514, 24)
 = \gcd(24, 10) \\
 = \gcd(10, 4) \\
 = \gcd(10, 2) \\
 = \gcd(2, 0) \\
\]
Lam\'e's work is generally considered the beginning of
computational complexity theory, which is the study of
resources needed (time or space) to execute an algorithm.
Another fascinating fact about Lam\'e's theorem is that historically
the above proof is the first ever \lq\lq use" of the Fibonacci sequence.

\begin{thm} \textnormal{(Lam\'e 1844)}
  Let $a > b > 0$ be integers.
  If the GCD computation of $a,b$ 
  using Euclidean algorithm results in $n$ steps:
  \[
  \gcd(a_{n+1}, b_{n+1})
  = \gcd(a_{n}, b_{n})
  = \cdots
  = \gcd(a_1, b_1), \,\,\, b_1 = 0
  \]
  where $(a_{n+1}, b_{n+1}) = (a, b)$, and $a_i > b_i$, then
  \begin{enumerate}[nosep,label=\textnormal{(\alph*)}]
  \item  $a \geq F_{n+2}$ and
    $b \geq F_{n+1}$, where $F_n$ are the Fibonacci numbers
    ($F_1 = 1, F_2 = 1, F_3 = 2, F_4 = 3, F_5 = 5$, etc.
    Note that the index starts with $1$.)
  \item $n$ is at most 5 times the number of digits in $b$.
  \end{enumerate}
\end{thm}
\proof
TODO
  
\begin{prop}
  The number of digits of a positive integer $b$ is
  \[
  \floor{\log_{10} b + 1}
  \]
\end{prop}
\proof
TODO

On analyzing the proof, the above is in fact true for any base $> 1$.
In other words the number of base--$B$ symbols to represent $b$ is
\[
\floor{\log_{B} b + 1}
\]
where $B > 1$.
For instance the number of bits needed to represent $b$ is
\[
\floor{\log_{2} b + 1}
\]
For instance, $b = 9_{10} = 1001_2$ which has 4 bits and
\[
\floor{\log_{2} 9 + 1}
= \floor{3.1699... + 1}
= \floor{4.1699...}
= 4
\]

\begin{prop}
  Let positive integer $b$ be written in base $B$ where $B > 1$ is an integer.
  Then the number of base-$B$ symbols used to represent $b$ is
  \[
  \floor{\log_{B} b + 1}
  \]
\end{prop}

\begin{ex}
  Leetcode 650.
  \\
  \url{https://leetcode.com/problems/2-keys-keyboard/}
  \\
  There is only one character \verb!'A'! on the screen of a notepad.
  You can perform one of two operations on this notepad for each step:
  \begin{enumerate}[nosep]
    \li Copy All:
    You can copy all the characters present on the screen (a partial copy is
    not allowed).
    \li Paste: You can paste the characters which are copied last time.
  \end{enumerate}
  Given an integer \verb!n!, return the minimum number of operations to get the character \verb!'A'! exactly \verb!n! times on the screen.
\end{ex}


\begin{ex}
  Leetcode 2447.
  \\
  \url{https://leetcode.com/problems/number-of-subarrays-with-gcd-equal-to-k}
  \\
  Given an integer array \verb!nums! and an integer \verb!k!,
  return the number of subarrays of \verb!nums! where the greatest common
  divisor
  of the subarray's elements is \verb!k!.
  A subarray is a contiguous non-empty sequence of elements within an array.
  The greatest common divisor of an array is the largest integer that evenly
  divides all the array elements.
\end{ex}

\begin{ex}
  Leetcode 1998
  \\
  \url{https://leetcode.com/problems/gcd-sort-of-an-array/}
  \\
  You are given an integer array \verb!nums!,
  and you can perform the following operation any number of times on
  \verb!nums!:
  \begin{enumerate}[nosep]
    \li Swap the positions of two elements \verb!nums[i]! and \verb!nums[j]! if
    \verb!gcd(nums[i], nums[j]) > 1! where \verb!gcd(nums[i], nums[j])! is the
    greatest common divisor of \verb!nums[i]! and \verb!nums[j]!.
  \end{enumerate}
  Return true if it is possible to sort \verb!nums! in non-decreasing order
  using the above swap method, or false otherwise.
\end{ex}
